R for Clinical Study Reports and Submission

### 2022-12-19 project-folder.Rmd

For a clinical trial project, it is also important to provide proper documentation for those help functions.
We use `roxygen2` to document functions.
For example, the header below defines each variable in `fmt_est`.
More details can be found in [Chapter 16 of the R Packages book](https://r-pkgs.org/man.html).

@@

The `roxygen2` documentation will be converted into
standard R documentation format, and saved as `.Rd` files in the `man/` folder.
This step is automatically handled by `devtools::document()`.

@@

The `man/` folder is used to save documentation automatically generated by `roxygen2`.
A typical workflow is to add `roxygen2` documentation before each function in the `R/` folder.
Then `devtools::document()` is used to generate all the documentation files in the `man/` folder.
More details can be found in [Chapter 16 of the R Packages book](https://r-pkgs.org/man.html).

@@

Reproducibility of analysis is one of the most important aspects of regulatory deliverables.
To ensure a successful reproduction, we need a controlled R environment,
including the control of the R version and the R package versions.
By using the R package folder structure and proper tools (e.g., `renv`, `packrat`),
we illustrate how to achieve reproducibility for R and R package versions.

@@

First, we introduce the control of the R version.
In the `esubdemo` project, a reproducible environment is created when you open
the `esubdemo.Rproj` from RStudio.
When we open the `esubdemo` project, RStudio will execute the R code in `.Rprofile` automatically.
So we can use `.Rprofile` to set up a reproducible environment.
More details can be found in <https://rstats.wtf/r-startup.html>.
After we open the `esubdemo` project, the code in `.Rprofile` will
automatically check the current R version is the same as we defined in `.Rprofile`.

@@

Secondly, we use `renv` to lock R package versions and save them in the `renv.lock` file.
`renv` provides a robust and stable approach to managing R package versions
for project-specific R packages.
An introduction of `renv` can be found on its
[website](https://rstudio.github.io/renv/articles/renv.html).

@@

The R code above in the `.Rprofile` initiates the `renv` running environment.
As a user, you can use `renv::init()`, `renv::snapshot()`, and `renv::restore()`
to initialize, save and restore R packages used for the current analysis project.

In the analysis project, the `renv` package will

@@

::: {.rmdnote data-latex=""}
The `renv.lock` file and `renv/` folder are only for project-specific R package.
A standard R package should not use `renv`.
:::

@@

- Define a snapshot date in `inst/startup.R`.
- Using `renv` to lock R versions within a project.

@@

- [`devtools`](https://devtools.r-lib.org/): make package development easier.
  + A good overview can be found in [Chapter 2 of the R Packages book](https://r-pkgs.org/Whole-game.html).
  + `devtools::load_all()`: load all functions in `R/` folder and running environment.
  + `devtools::document()`: automatically create documentation using `roxygen2`.
  + `devtools::check()`: automatically perform compliance check as an R package.
  + `devtools::build_site()`: automatically run analysis scripts in batch and create a `pkgdown` website.
- [`usethis`](https://usethis.r-lib.org/): automates repetitive tasks that arise during project setup and development.
- [`testthat`](https://testthat.r-lib.org/): streamline testing code.
  + A discussion of using the `testthat` for an A&R project can be found in (@madhutest).
- [`pkgdown`](https://pkgdown.r-lib.org/): generate static HTML documentation website for an R package
  + It also allows you to run all analysis code in batch.

>>>>

For a clinical trial project, it is also important to provide proper documentation for those help functions.
We use roxygen2 package to document functions.
For example, the header below defines each variable in `fmt_est`.
More details can be found in [Chapter 16 of the R Packages book](https://r-pkgs.org/man.html).

@@

The roxygen2 documentation will be converted into
standard R documentation format, and saved as `.Rd` files in the `man/` folder.
This step is automatically handled by `devtools::document()`.

@@

The `man/` folder is used to save documentation automatically generated by roxygen2.
A typical workflow is to add roxygen2 documentation before each function in the `R/` folder.
Then `devtools::document()` is used to generate all the documentation files in the `man/` folder.
More details can be found in [Chapter 16 of the R Packages book](https://r-pkgs.org/man.html).

@@

Reproducibility of analysis is one of the most important aspects of regulatory deliverables.
To ensure a successful reproduction, we need a controlled R environment,
including the control of the R version and the R package versions.
By using the R package folder structure and proper tools (e.g., renv, packrat),
we illustrate how to achieve reproducibility for R and R package versions.

@@

First, we introduce the control of the R version.
In the esubdemo project, a reproducible environment is created when you open
the `esubdemo.Rproj` from RStudio.
When we open the `esubdemo` project, RStudio will execute the R code in `.Rprofile` automatically.
So we can use `.Rprofile` to set up a reproducible environment.
More details can be found in <https://rstats.wtf/r-startup.html>.
After we open the `esubdemo` project, the code in `.Rprofile` will
automatically check the current R version is the same as we defined in `.Rprofile`.

@@

Secondly, we use renv to lock R package versions and save them in the `renv.lock` file.
renv provides a robust and stable approach to managing R package versions
for project-specific R packages.
An introduction of renv can be found on its
[website](https://rstudio.github.io/renv/articles/renv.html).

@@

The R code above in the `.Rprofile` initiates the renv running environment.
As a user, you can use `renv::init()`, `renv::snapshot()`, and `renv::restore()`
to initialize, save and restore R packages used for the current analysis project.

In the analysis project, the renv package will

@@

::: {.rmdnote data-latex=""}
The `renv.lock` file and `renv/` folder are only for project-specific R package.
A standard R package should not use renv.
:::

@@

- Define a snapshot date in `inst/startup.R`.
- Using renv to lock R versions within a project.

@@

- [devtools](https://devtools.r-lib.org/): make package development easier.
  + A good overview can be found in [Chapter 2 of the R Packages book](https://r-pkgs.org/Whole-game.html).
  + `devtools::load_all()`: load all functions in `R/` folder and running environment.
  + `devtools::document()`: automatically create documentation using roxygen2.
  + `devtools::check()`: automatically perform compliance check as an R package.
  + `devtools::build_site()`: automatically run analysis scripts in batch and create a pkgdown website.
- [usethis](https://usethis.r-lib.org/): automates repetitive tasks that arise during project setup and development.
- [testthat](https://testthat.r-lib.org/): streamline testing code.
  + A discussion of using the testthat for an A&R project can be found in (@madhutest).
- [pkgdown](https://pkgdown.r-lib.org/): generate static HTML documentation website for an R package
  + It also allows you to run all analysis code in batch.

### 2022-12-19 project-management.Rmd

::: {.rmdnote}
After project initiation, modifying `.Rprofile` will be a risk of reproducibility
and should be handled carefully if necessary.
:::

@@

In an R package development, the validation or testing is completed under the `test/` folder.
The testthat R package can be used to streamline the validation process.
More details of the `testthat` package for R package validation can be found in [Chapter 12 of the R package book](https://r-pkgs.org/tests.html).

@@

The project lead can also use the R package `pkgdown` to build a complete website for a project-specific R package.
The `pkgdown` website is a convenient way to run all analyses in batch and
integrate outputs in a website, which comprehensively covers
project-specific R functions, TLF generation programs, outputs and validation tracking information, etc.
For example, in the `esubdemo` project, we created the `pkgdown` website at <https://elong0527.github.io/esubdemo/>.

Many of the tasks in SDLC can be completed automatically.
An organization can leverage CI/CD workflow to automatically enable those tasks,
such as running testing cases and creating a `pkgdown` website.
For example, in the `esubdemo` project, we set up
[GitHub Actions](https://github.com/elong0527/esubdemo/actions)
for it. This can be done by using `usethis::use_github_action()`.

>>>>

::: {.rmdnote data-latex=""}
After project initiation, modifying `.Rprofile` will be a risk of reproducibility
and should be handled carefully if necessary.
:::

@@

In an R package development, the validation or testing is completed under the `test/` folder.
The testthat R package can be used to streamline the validation process.
More details of the testthat package for R package validation can be found in [Chapter 12 of the R package book](https://r-pkgs.org/testing-basics.html).

@@

The project lead can also use the R package pkgdown to build a complete website for a project-specific R package.
The pkgdown website is a convenient way to run all analyses in batch and
integrate outputs in a website, which comprehensively covers
project-specific R functions, TLF generation programs, outputs and validation tracking information, etc.
For example, in the `esubdemo` project, we created the pkgdown website at <https://elong0527.github.io/esubdemo/>.

Many of the tasks in SDLC can be completed automatically.
An organization can leverage CI/CD workflow to automatically enable those tasks,
such as running testing cases and creating a pkgdown website.
For example, in the `esubdemo` project, we set up
[GitHub Actions](https://github.com/elong0527/esubdemo/actions)
for it. This can be done by using `usethis::use_github_action()`.

### 2022-12-19 slides/r4csr-rstudio.Rmd

- As statisticians, we use `tidyverse`, `r2rtf`,
and internal tools to define the mock-up table, listing and figure (TLFs) for statistical analysis of a clinical trial.
  - More than 100 TLFs for efficacy and safety of a drug or vaccine.

- As programmers, we use `tidyverse`, `r2rtf`, and internal tools to develop and/or validate analysis results based on mock-up TLFs.

>>>>

- As statisticians, we use tidyverse, r2rtf,
and internal tools to define the mock-up table, listing and figure (TLFs) for statistical analysis of a clinical trial.
  - More than 100 TLFs for efficacy and safety of a drug or vaccine.

- As programmers, we use tidyverse, r2rtf, and internal tools to develop and/or validate analysis results based on mock-up TLFs.

### 2022-12-19 tlf-ae-spec.Rmd

To obtain the nested layout, we use the `page_by` argument in the
`rtf_body` function. By defining `page_by="AESOC"`, `r2rtf` recognizes
the variable as a group indicator.

@@

More discussion on `page_by`, `group_by` and `subline_by`
features can be found on the
[`r2rtf` package website](https://merck.github.io/r2rtf/articles/example-sublineby-pageby-groupby.html).

@@

- Step 1: Read data (i.e., `adae` and `adsl`) into R.
- Step 2: Count the number of participants by SOC and treatment arm
  (rows with bold text) and save into `t1`.
- Step 3: Count the number of participants in each AE term by SOC, AE
  term, and treatment arm (rows without bold text) and save into `t2`.
- Step 4: Bind `t1` and `t2` by row into `t_ae`.
- Step 5: Count the number of participants in each arm as `t_pop`.
- Step 6: Row-wise combine `t_pop` and `t_ae` into `tbl_ae_spec`.
- Step 7: Format the output by using `r2rtf`.

>>>>

To obtain the nested layout, we use the `page_by` argument in the
`rtf_body` function. By defining `page_by="AESOC"`, r2rtf recognizes
the variable as a group indicator.

@@

More discussion on `page_by`, `group_by` and `subline_by`
features can be found on the
[r2rtf package website](https://merck.github.io/r2rtf/articles/example-sublineby-pageby-groupby.html).

@@

- Step 1: Read data (i.e., `adae` and `adsl`) into R.
- Step 2: Count the number of participants by SOC and treatment arm
  (rows with bold text) and save into `t1`.
- Step 3: Count the number of participants in each AE term by SOC, AE
  term, and treatment arm (rows without bold text) and save into `t2`.
- Step 4: Bind `t1` and `t2` by row into `t_ae`.
- Step 5: Count the number of participants in each arm as `t_pop`.
- Step 6: Row-wise combine `t_pop` and `t_ae` into `tbl_ae_spec`.
- Step 7: Format the output by using r2rtf.

### 2022-12-19 tlf-overview.Rmd

::: {.rmdnote data-latex=""}
There are other R packages to create TLFs in ASCII, RTF and Word format.
For example, `rtables`, `huxtable`, `pharmaRTF`, `gt`, `officer`, `flextable` etc.
Here we focus on `r2rtf` to illustrate the concept.  
Readers are encouraged to explore other R packages to find the proper tools to fit your purpose. 
:::

@@

`tidyverse` is a collection of R packages to simplify the workflow to manipulate,
visualize and analyze data in R.
Those R packages share
[the tidy tools manifesto](https://cran.r-project.org/web/packages/tidyverse/vignettes/manifesto.html)
and are easy to use for interactive data analysis.

RStudio provided outstanding [cheatsheets](https://www.rstudio.com/resources/cheatsheets/)
and [tutorials](https://github.com/rstudio-education/remaster-the-tidyverse) for `tidyverse`.

There are also books to introduce `tidyverse`.
We assume the reader have experience in using `tidyverse` in this book.

@@

`r2rtf` is an R package to create production-ready tables and figures in RTF format.
This R package is designed to

- provide simple "verb" functions that correspond to each component of a table,
  to help you translate a data frame to a table in an RTF file;
- enable pipes (`%>%`);
- focus on the **table format** only.
  Data manipulation and analysis shall be handled by other R packages (e.g., `tidyverse`).

@@

We provide a brief introduction of `r2rtf` and show how to transfer
data frames into table, listing, and figures (TLFs).

Other extended examples and features are covered on the
[`r2rtf` package website](https://merck.github.io/r2rtf/articles/index.html).

To explore the basic RTF generation verbs in `r2rtf`,
we will use the dataset `r2rtf_adae` saved in the `r2rtf` package.
This dataset contains adverse events (AEs) information from a clinical trial.

@@

`dplyr` and `tidyr` packages within `tidyverse` are used
for data manipulation to create a data frame
that contains all the information we want to add in an RTF table.

@@

`r2rtf` aims to provide one function for each type of table layout.
Commonly used verbs include:

@@

`r2rtf` supports 26 different border types. The details can be found on
the [r2rtf package website](https://merck.github.io/r2rtf/articles/rtf-row.html#border-type).

@@

In the `r2rtf` R package [get started](https://merck.github.io/r2rtf/articles/r2rtf.html) page,
there are more examples to illustrate how to customize

>>>>

::: {.rmdnote data-latex=""}
There are other R packages to create TLFs in ASCII, RTF and Word format.
For example, rtables, huxtable, pharmaRTF, gt, officer, flextable etc.
Here we focus on r2rtf to illustrate the concept.  
Readers are encouraged to explore other R packages to find the proper tools to fit your purpose. 
:::

@@

tidyverse is a collection of R packages to simplify the workflow to manipulate,
visualize and analyze data in R.
Those R packages share
[the tidy tools manifesto](https://cran.r-project.org/web/packages/tidyverse/vignettes/manifesto.html)
and are easy to use for interactive data analysis.

RStudio provided outstanding [cheatsheets](https://www.rstudio.com/resources/cheatsheets/)
and [tutorials](https://github.com/rstudio-education/remaster-the-tidyverse) for tidyverse.

There are also books to introduce tidyverse.
We assume the reader have experience in using tidyverse in this book.

@@

r2rtf is an R package to create production-ready tables and figures in RTF format.
This R package is designed to

- provide simple "verb" functions that correspond to each component of a table,
  to help you translate a data frame to a table in an RTF file;
- enable pipes (`%>%`);
- focus on the **table format** only.
  Data manipulation and analysis shall be handled by other R packages (e.g., tidyverse).

@@

We provide a brief introduction of r2rtf and show how to transfer
data frames into table, listing, and figures (TLFs).

Other extended examples and features are covered on the
[r2rtf package website](https://merck.github.io/r2rtf/articles/index.html).

To explore the basic RTF generation verbs in r2rtf,
we will use the dataset `r2rtf_adae` saved in the r2rtf package.
This dataset contains adverse events (AEs) information from a clinical trial.

@@

dplyr and tidyr packages within tidyverse are used
for data manipulation to create a data frame
that contains all the information we want to add in an RTF table.

@@

r2rtf aims to provide one function for each type of table layout.
Commonly used verbs include:

@@

r2rtf supports 26 different border types. The details can be found on
the [r2rtf package website](https://merck.github.io/r2rtf/articles/rtf-row.html#border-type).

@@

In the r2rtf R package [get started](https://merck.github.io/r2rtf/articles/r2rtf.html) page,
there are more examples to illustrate how to customize
